---
title: "Hands-on Exercise 9:Geographically Weighted Predictive Models"
format:
  html:
    toc: true
execute: 
  eval: false
  warning: false
  freeze: true
  fig-retine: 3
  date: "March 18,2024"
date-modified: "last-modified"
---

# 9 Overview

Predictive modelling uses statistical learning or machine learning techniques to predict outcomes. By and large, the event one wants to predict is in the future. However, a set of known outcome and predictors (also known as variables) will be used to calibrate the predictive models.

Geospatial predictive modelling is conceptually rooted in the principle that the occurrences of events being modeled are limited in distribution. When geographically referenced data are used, occurrences of events are neither uniform nor random in distribution over space. There are geospatial factors (infrastructure, sociocultural, topographic, etc.) that constrain and influence where the locations of events occur. Geospatial predictive modeling attempts to describe those constraints and influences by spatially correlating occurrences of historical geospatial locations with environmental factors that represent those constraints and influences.

# 9.1 Data

-   **Aspatial dataset**:

    -   HDB Resale data: a list of HDB resale transacted prices in Singapore from Jan 2017 onwards. It is in csv format which can be downloaded from Data.gov.sg.

-   **Geospatial dataset**:

    -   *MP14_SUBZONE_WEB_PL*: a polygon feature data providing information of URA 2014 Master Plan Planning Subzone boundary data. It is in ESRI shapefile format. This data set was also downloaded from Data.gov.sg

-   **Locational factors with geographic coordinates**:

    -   Downloaded from **Data.gov.sg**.

        -   **Eldercare** data is a list of eldercare in Singapore. It is in shapefile format.

        -   **Hawker Centre** data is a list of hawker centres in Singapore. It is in geojson format.

        -   **Parks** data is a list of parks in Singapore. It is in geojson format.

        -   **Supermarket** data is a list of supermarkets in Singapore. It is in geojson format.

        -   **CHAS clinics** data is a list of CHAS clinics in Singapore. It is in geojson format.

        -   **Childcare service** data is a list of childcare services in Singapore. It is in geojson format.

        -   **Kindergartens** data is a list of kindergartens in Singapore. It is in geojson format.

    -   Downloaded from **Datamall.lta.gov.sg**.

        -   **MRT** data is a list of MRT/LRT stations in Singapore with the station names and codes. It is in shapefile format.

        -   **Bus stops** data is a list of bus stops in Singapore. It is in shapefile format.

-   **Locational factors without geographic coordinates**:

    -   Downloaded from **Data.gov.sg**.

        -   **Primary school** data is extracted from the list on General information of schools from data.gov portal. It is in csv format.

    -   Retrieved/Scraped from **other sources**

        -   **CBD** coordinates obtained from Google.

        -   **Shopping malls** data is a list of Shopping malls in Singapore obtained from [Wikipedia](https://en.wikipedia.org/wiki/List_of_shopping_malls_in_Singapore).

        -   **Good primary schools** is a list of primary schools that are ordered in ranking in terms of popularity and this can be found at [Local Salary Forum](https://www.salary.sg/2021/best-primary-schools-2021-by-popularity).

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
              tmap, rsample, Metrics, tidyverse)
```

## **9.2 Preparing Data**

### **9.2.1 Reading Data File to rds**

```{r}
mdata <- read_rds("data/aspatial/mdata.rds")
```

### **9.2.2 Data Sampling**

```{r}
set.seed(1234)
resale_split <- initial_split(mdata, 
                              prop = 6.5/10,)
train_data <- training(resale_split)
test_data <- testing(resale_split)
```

```{r}
write_rds(train_data, "data/model/train_data.rds")
write_rds(test_data, "data/model/test_data.rds")
```

## **9.3 Computing Correlation Matrix**

```{r}
mdata_nogeo <- mdata %>%
  st_drop_geometry()
corrplot::corrplot(cor(mdata_nogeo[, 2:17]), 
                   diag = FALSE, 
                   order = "AOE",
                   tl.pos = "td", 
                   tl.cex = 0.5, 
                   method = "number", 
                   type = "upper")
```

## **9.4 Retriving the Stored Data**

```{r}
train_data <- read_rds("data/model/train_data.rds")
test_data <- read_rds("data/model/test_data.rds")
```

## **9.5 Building a Non-spatial Multiple Linear Regression**

```{r}
price_mlr <- lm(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                data=train_data)
summary(price_mlr)
```

```{r}
write_rds(price_mlr, "model/price_mlr.rds" ) 
```

## **9.6 GWR Predictive Method**

### **9.6.1 Converting the sf data.frame to SpatialPointDataFrame**

```{r}
train_data_sp <- as_Spatial(train_data)
train_data_sp
```

### **9.6.2 Computing Adaptive Bandwidth**

```{r}
bw_adaptive <- bw.gwr(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                  data=train_data_sp,
                  approach="CV",
                  kernel="gaussian",
                  adaptive=TRUE,
                  longlat=FALSE)
```

```{r}
write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```

### **9.6.3 Constructing the Adaptive Bandwidth GWR Model**

```{r}
bw_adaptive <- read_rds("data/model/bw_adaptive.rds")
```

```{r}
gwr_adaptive <- gwr.basic(formula = resale_price ~
                            floor_area_sqm + storey_order +
                            remaining_lease_mths + PROX_CBD + 
                            PROX_ELDERLYCARE + PROX_HAWKER +
                            PROX_MRT + PROX_PARK + PROX_MALL + 
                            PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                            WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                            WITHIN_1KM_PRISCH,
                          data=train_data_sp,
                          bw=bw_adaptive, 
                          kernel = 'gaussian', 
                          adaptive=TRUE,
                          longlat = FALSE)
```

```{r}
write_rds(gwr_adaptive, "data/model/gwr_adaptive.rds")
```

### **9.6.4 Retrieve GWR Output Object**

```{r}
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
```

```{r}
gwr_adaptive
```

### **9.6.5 Converting the Test Data from Special Feature Data Frame to Spatial Point Data Frame**

```{r}
test_data_sp <- test_data %>%
  as_Spatial()
test_data_sp
```

### **9.6.6 Computing Adaptive Bandwidth for the Test Data**

```{r}
gwr_bw_test_adaptive <- bw.gwr(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                  data=test_data_sp,
                  approach="CV",
                  kernel="gaussian",
                  adaptive=TRUE,
                  longlat=FALSE)
```

### **9.6.7 Computing Predicted Values of the Test Data**

```{r}
gwr_pred <- gwr.predict(formula = resale_price ~
                          floor_area_sqm + storey_order +
                          remaining_lease_mths + PROX_CBD + 
                          PROX_ELDERLYCARE + PROX_HAWKER + 
                          PROX_MRT + PROX_PARK + PROX_MALL + 
                          PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                          WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
                          WITHIN_1KM_PRISCH, 
                        data=train_data_sp, 
                        predictdata = test_data_sp, 
                        bw=40, 
                        kernel = 'gaussian', 
                        adaptive=TRUE, 
                        longlat = FALSE)
```

## **9.7 Preparing Coordinates Data**

### **9.7.1 Extracting Coordinates Data**

```{r}
coords <- st_coordinates(mdata)
coords_train <- st_coordinates(train_data)
coords_test <- st_coordinates(test_data)
```

```{r}
coords_train <- write_rds(coords_train, "data/model/coords_train.rds" )
coords_test <- write_rds(coords_test, "data/model/coords_test.rds" )
```

### **9.7.2 Dropping Geometry Field**

```{r}
train_data <- train_data %>% 
  st_drop_geometry()
```

## **9.8 Calibrating Random Forest Model**

```{r}
set.seed(1234)
rf <- ranger(resale_price ~ floor_area_sqm + storey_order + 
               remaining_lease_mths + PROX_CBD + PROX_ELDERLYCARE + 
               PROX_HAWKER + PROX_MRT + PROX_PARK + PROX_MALL + 
               PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
               WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
               WITHIN_1KM_PRISCH,
             data=train_data)
rf
```

```{r}
write_rds(rf, "data/model/rf.rds")
```

```{r}
rf <- read_rds("data/model/rf.rds")
rf
```

## **9.9 Calibrating Geographical Random Forest Model**

### **9.9.1 Calibrating using Training Data**

```{r}
set.seed(1234)
gwRF_adaptive <- grf(formula = resale_price ~ floor_area_sqm + storey_order +
                       remaining_lease_mths + PROX_CBD + PROX_ELDERLYCARE +
                       PROX_HAWKER + PROX_MRT + PROX_PARK + PROX_MALL +
                       PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                       WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                       WITHIN_1KM_PRISCH,
                     dframe=train_data, 
                     bw=55,
                     kernel="adaptive",
                     coords=coords_train)
```

```{r}
write_rds(gwRF_adaptive, "data/model/gwRF_adaptive.rds")
```

```{r}
gwRF_adaptive <- read_rds("data/model/gwRF_adaptive.rds")
```

### **9.9.2 Predicting by using Test Data**

#### 9.9.2.1 Preparing the Test Data

```{r}
test_data <- cbind(test_data, coords_test) %>%
  st_drop_geometry()
```

#### 9.9.2.2 Predicting with Test Data

```{r}
gwRF_pred <- predict.grf(gwRF_adaptive, 
                           test_data, 
                           x.var.name="X",
                           y.var.name="Y", 
                           local.w=1,
                           global.w=0)
```

```{r}
GRF_pred <- write_rds(gwRF_pred, "data/model/GRF_pred.rds")
```

#### 9.9.2.3 Converting the Predicting Output into a Data Frame

```{r}
GRF_pred <- read_rds("data/model/GRF_pred.rds")
GRF_pred_df <- as.data.frame(GRF_pred)
```

```{r}
test_data_p <- cbind(test_data, GRF_pred_df)
```

```{r}
write_rds(test_data_p, "data/model/test_data_p.rds")
```

### **9.9.3 Calculating Root Mean Square Error**

```{r}
rmse(test_data_p$resale_price, 
     test_data_p$GRF_pred)
```

### **9.9.4 Visualising the Predicted Values**

```{r}
ggplot(data = test_data_p,
       aes(x = GRF_pred,
           y = resale_price)) +
  geom_point()
```
